---
date: "`r Sys.Date()`"
output: html_document
---

# **Trajectory analysis**

Trajectory analysis is a powerful tool in single-cell RNA-seq allowing us to track how cells transition between different states over time, rather than simply classifying them into discrete clusters. This approach is particularly useful for studying dynamic processes such as cellular differentiation, cell cycle progression, or disease progression. One of the key advantages of trajectory analysis is the ability to identify genes or (in the case of long read data) isoforms whose expression patterns are correlated with pseudotime

In our example dataset we are interested in looking at genes or isoforms that change as a function of pseudotime as this will highlight impor features that control stem cell to neuronal differentiation. In the bellow code block is the `perform_trajectory_analysis` function which uses monocle3 [@trapnell2014] to learn the trajectory and order the cells by pseudotime. We also provide an optional mode to to identify gene or isoforms that change with pseudotime. [^09_trajectory-1]

[^09_trajectory-1]: It’s important to note that analyzing features in relation to pseudotime can be computationally intensive Users can choose to plot the trajectory without the added computational load of ordering features by pseudotime, making the analysis more flexible. The parameter `number_of_features` can be used to specify the number of highly variable features to test.

    .

```{r traj_fucntion, echo=TRUE, include=TRUE}

    perform_trajectory_analysis <- function(seurat_obj, assay, prefix, cluster_name="harm_cluster", 
                                            root_cells = NULL, cores = 8, order_features_by_pseudotime = FALSE, number_of_features = 500) {
      library(monocle3)
      library(SeuratWrappers)
      
      # Step 1: Subset and Remove RNA Assay
      DefaultAssay(seurat_obj) <- assay
      seurat_obj <- JoinLayers(seurat_obj)
      
      # Step 2: Convert to Monocle cell_data_set object
      cds <- as.cell_data_set(seurat_obj)
      
      # Add gene_short_name to feature data (gene metadata)
      fData(cds)$gene_short_name <- rownames(fData(cds))
      
      # Estimate size factors
      cds <- estimate_size_factors(cds)
      
      # Step 3: Assign cluster info and UMAP coordinates
      # Partitions
      recreate_partition <- as.factor(rep(1, length(colnames(seurat_obj))))
      names(recreate_partition) <- colnames(seurat_obj)
      cds@clusters$UMAP$partitions <- recreate_partition
      
      # Clustering information from Seurat
      cds@clusters$UMAP$clusters <- seurat_obj@meta.data[["orig.ident"]]
      
      # UMAP embeddings from Seurat
      cds@int_colData@listData$reducedDims$UMAP <- seurat_obj@reductions$umap.harm@cell.embeddings
      
      # Plot before trajectory
      cluster_before_trajectory <- plot_cells(cds, color_cells_by = 'orig.ident') +
        theme(legend.position = "right")
      
      # Save plot
      print(cluster_before_trajectory)
      
      # Step 4: Learn trajectory graph
      cds <- learn_graph(cds, use_partition = FALSE, close_loop = FALSE)
      
      # Plot learned trajectory
      plot_trajectory <- plot_cells(cds, color_cells_by = cluster_name) +
        theme(legend.position = "right")
      
      # Step 5: Order cells in pseudotime
      cds <- order_cells(cds, reduction_method = 'UMAP', root_cells = root_cells)
      
      # Plot pseudotime
      p_pseudotime <- plot_cells(cds, color_cells_by = 'pseudotime') +
        theme(legend.position = "right")
      
      # plot
      print(plot_trajectory | p_pseudotime)
      
      # Save pseudotime to Seurat object
      seurat_obj$pseudotime <- pseudotime(cds)
      
      if (order_features_by_pseudotime == TRUE) {
      # Step 6: Finding genes that change as a function of pseudotime
      # Find the X most variable genes
      seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", nfeatures = number_of_features)

      # Extract the top n variable gene names
      top_variable_genes <- VariableFeatures(seurat_obj)
      
      cds_subset <- cds[top_variable_genes, ]
      
      deg_results <- graph_test(cds_subset, neighbor_graph = 'principal_graph', cores = cores) 
      
      # Filter for significant genes
      MI_sig <- deg_results %>%
        filter(q_value < 0.01, morans_I > 0.5, status == 'OK') %>%
        arrange(q_value)
      
      # Save significant genes
      write.csv(MI_sig, paste0(prefix, "_sig_genes_pseudotime.csv"))
      
      # Plot top genes
      top_genes <- MI_sig %>% 
        arrange(desc(morans_I)) %>% 
        head(16)
      
      #print plot
      print(FeaturePlot(seurat_obj, features = top_genes$gene_short_name, reduction ="umap.harm"))

      # Another plot for top 16 fetautes DE acrose pseudotime 
      my_genes <- row.names(subset(fData(cds_subset), gene_short_name %in% top_genes$gene_short_name)) 
      #my_genes <- row.names(subset(fData(cds), gene_short_name %in% c("DLK1","GRIA2","GRIN2B", "VIM", "PKM", "GAD1", "DCX", "POU5F1", "HOPX"))) 
      
      cds_top_genes <- cds_subset[my_genes,]
      
      monocle_plot <- monocle3::plot_genes_in_pseudotime(cds = cds_top_genes, min_expr = 0.5, cell_size = 0.5,
                                         nrow = NULL, ncol = 4, panel_order = NULL,
                                         label_by_short_name = TRUE, color_cells_by = cluster_name)
      
      print(monocle_plot)

        }
      
      # Save the Seurat object with pseudotime info
      #saveRDS(seurat_obj, file = paste0(prefix, "_seurat_with_pseudotime.rds"))

    }
```

From this analysis, we observe a well-defined trajectory that transitions from stem cells through immature neurons (likely neural progenitor cells) to radial glial cells, and finally toward more mature neuronal subtypes. This trajectory aligns well with known biological processes, supporting its validity.

Given this biologically meaningful trajectory, we examined genes that exhibit expression changes along pseudotime. Below are 16 top hits, identified based on the number of features tested, showing significant correlations between expression and trajectory progression. These genes may serve as markers for specific cell clusters or be differentially expressed based on other testing methods. We have visualized these results using both a feature plot and a Monocle3 dot plot.

Notes:

1.   Add labelles for figures
2.  correct figure size.
3.  Add reduction param
4.   work out the empty space

```{r traj_genes, echo=TRUE, fig.height=6, fig.width=10, message=FALSE, warning=FALSE}
perform_trajectory_analysis(seurat_obj = obj, "RNA", prefix = "Multi-sample_tutorial",
                            cluster_name="sctype_db",  root_cells = NULL, cores = 8, order_features_by_pseudotime = TRUE, number_of_features = 500)

### need to add in output path
```

The same analysis can be performed at the isoform level by setting `assay = "iso"`. This allows for the identification of differentially expressed isoforms along the trajectory. Users can also plot trajectories based on isoform-level clustering. If choosing this approach, ensure that the isoform-level data is properly integrated, and modify the function to specify the appropriate dimensional reduction method for trajectory learning.

```{r traj_iso, echo=TRUE, fig.height=6, fig.width=10, message=FALSE, warning=FALSE}
perform_trajectory_analysis(seurat_obj = obj, assay="iso", prefix = "iso_Multi-sample_tutorial",
                            cluster_name="sctype_db",  root_cells = NULL, cores = 8, order_features_by_pseudotime = TRUE, number_of_features = 1000)

```
